Lord Kelvin's debate over the age of the Earth: an early attempt of interdisciplinary science

### 2023-08-15 ms.tex
wordsmith intro

Many of the big unsolved problems in science today will require an interdisciplinary approach to solve. For example, the unsolved problem of the origin of life on Earth can only be tackled by drawing upon many different fields of research \citep{Hays_2015}. Geologic expertise is required to understand the earliest fossil records of life, organic chemists are needed to synthesize early life-like replicating molecules, and atmospheric modelers are essential to estimate the plausible conditions on the early Earth under which life began. Another problem that requires broad expertise is projecting the impact of anthropogenic climate change on society. Any forecast will involve atmospheric scientists, glaciologists, economists, and ecologists among many other areas of expertise. Given that important unsolved problems rely on an interdisciplinary approach, it is essential scientists embedded in different departments and academic cultures learn to work together effectively.

One of the earliest attempts of interdisciplinary science is the debate over the age of the Earth in the late 1800s. The physicist Lord Kelvin used a thermodynamic model of Earth's heat flow to estimate Earth was only $\sim 100$ million years old. He brought this calculation to the attention of geologists, who generally assumed the Earth was indefinitely old based on the paradigm of uniformitarianism. A relatively young Earth also troubled Darwin, who worried his theory of biological evolution needed more time than Kelvin allowed. The ensuing debate, involving physics, geology and biology was one of the earliest attempts of interdisciplinary science. However, the scientific discourse did not proceed smoothly. Geologists and physicists talked past each other, and neither discipline was capable of entertaining evidence outside their domain of expertise.

>>>>

Many of the big unsolved problems in science today will require an interdisciplinary approach to answer. For example, the unsolved problem of the origin of life on Earth can only be tackled using many different fields of research \citep{Hays_2015}. Geologic expertise is required to understand the earliest fossil records of life, organic chemists are needed to synthesize life-like replicating molecules, and atmospheric modelers are essential to estimate the plausible conditions on the early Earth under which life began. Another problem that requires broad expertise is projecting the impact of anthropogenic climate change on society. Any forecast will involve atmospheric scientists, glaciologists, economists, and ecologists among many other areas of knowledge. Given that important unanswered problems rely on an interdisciplinary approach, it is essential that scientists work together effectively even if they are embedded in different departments with ranging academic cultures.

One of the earliest attempts of interdisciplinary science is the debate over the age of the Earth in the late 1800s. The physicist Lord Kelvin used a thermodynamic model of heat flow in the crust to estimate Earth's age to be only $\sim 100$ million years old. He brought this calculation to the attention of geologists, who generally assumed the Earth was indefinitely old based on the paradigm of uniformitarianism. A relatively young Earth also troubled Darwin, who worried his theory of biological evolution needed more time than Kelvin allowed. The ensuing debate, involving physics, geology and biology was one of the earliest attempts of interdisciplinary science. However, the scientific discourse did not proceed smoothly. Geologists and physicists talked past each other, and neither discipline was capable of entertaining evidence outside their domain of expertise.
